\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{microtype}
\usepackage{authblk}
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}

\setlength{\emergencystretch}{2em}

\title{\textbf{$\sqrt{S}$ KV-Cache Residency for Long-Context LLM Decode:\\
Real Costs, Real Correctness, and a Real Capability Unlock\\
on Memory-Constrained Hardware}}

\author[1]{Jun Kawasaki}
\affil[1]{com-junkawasaki / etzhayyim / gftdcojp}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
Autoregressive decode stores a key--value (KV) cache that grows linearly
with sequence length $S$; on memory-constrained hardware this cache, not
model weights, is what prevents a model from reaching the long context it
was trained for. We study \emph{$\sqrt{S}$ residency}: partition the
sequence into blocks of stride $b\approx\sqrt{S\log S}$, keep only boundary
checkpoints plus one active block resident, and page the rest from host
memory or disk, combined with online-softmax so attention stays exact. The
checkpoint-and-recompute pattern itself is classical --- it is the
reverse-mode-differentiation checkpointing of
Griewank--Walther~\cite{griewank2000revolve} and
Chen~et~al.~\cite{chen2016checkpoint}; we borrow Williams'
(STOC~2025)~\cite{williams2025} improved stride formula as a
\emph{principled default} for the block size, not as a claim that
Transformer decode is a multitape Turing machine or that this technique
depends on his proof. We validate the full chain empirically, not just the
storage-ratio headline: (1)~\textbf{correctness end-to-end} --- a real
\texttt{Cache} implementation fused into a real multi-token generation loop
produces byte-identical output to standard full-KV decode, not just a
single-tensor numerical check; (2)~\textbf{honest cost} --- re-paging the
non-resident set every decode step is $2.2$--$3.7\times$ slower, measured
two independent ways on NVIDIA A100, worsening with $S$; (3)~a
\textbf{competitive comparison} --- at the $\sim\!90\%$ compression ratio
this method's own headline claims, three lossy eviction baselines
(StreamingLLM~\cite{xiao2023streamingllm}, a norm-based scorer, and
SnapKV~\cite{li2024snapkv}) score $0\%$ accuracy on a needle-retrieval task
where $\sqrt{S}$ residency is exact by construction; (4)~a
\textbf{hardware-specific finding} on Apple Silicon's unified memory: naive
host offloading does \emph{not} free usable memory there (measured, not
assumed), while memory-mapped paging does, confirmed up to $6.3\,\mathrm{GB}$
of non-resident data at $0.0\,\mathrm{MB}$ RSS cost; and (5)~a
\textbf{real capability unlock}: on a real 9B model with a $1.05\times10^6$
token native context, a $16\,\mathrm{GB}$ Mac can reach only $6.1\%$ of that
context without this technique. We also report negative results
transparently: whole-prefix recompute is not a peak-memory win, and neither
the PyTorch/HuggingFace nor the MLX reference implementation of DeepSeek-V2's
Multi-head Latent Attention~\cite{deepseekv2} actually caches its compressed
latent, so true composability with an \emph{absorbed}-cache MLA
implementation remains open. Code and JSON benchmarks:
\texttt{github.com/gftdcojp/cloud-murakumo}.
\end{abstract}

\section{Introduction}

Autoregressive Transformers store key/value tensors for every past token.
For sequence length $S$, layers $L$, KV heads $H_{\mathrm{kv}}$, and head
dimension $d$, the cache occupies
\begin{equation}
  \mathrm{KVBytes}(S)
  \;=\;
  2\, L\, H_{\mathrm{kv}}\, d\, S\, \mathrm{sizeof}(\mathrm{dtype}),
\end{equation}
i.e.\ $\Theta(S)$ for fixed model shape. On cloud GPUs this competes with
batching headroom; on memory-constrained hardware --- a laptop, a phone, an
embedded device --- it can be the difference between a model running at all
and not. \textbf{This paper's framing is capability, not speed}: we ask
whether a model can be made to reach a context length it otherwise could
not on a fixed memory budget, accepting a decode-latency cost as the price.

Reducing peak memory by trading it for recomputation is a decades-old idea
in the reverse-mode automatic differentiation literature: divide a
computation into $\Theta(\sqrt{t})$ segments, keep $\Theta(\sqrt{t})$
checkpoints, and recompute the rest on demand, giving
$\Theta(\sqrt{t})$ space at $O(1)$--$O(\log t)$ recomputation
overhead~\cite{griewank2000revolve}. Chen et al.\ brought the same idea to
neural network training memory~\cite{chen2016checkpoint}. Independently,
Williams (STOC~2025) improved the generic simulation of time by space for
multitape Turing machines from Hopcroft--Paul--Valiant's
$O(t/\log t)$~\cite{hpv1977} to $O(\sqrt{t\log t})$, via a reduction to Tree
Evaluation and the space-efficient algorithm of
Cook--Mertz~\cite{cookmertz2024}. The two lines of work converge on the same
block-size arithmetic; we adopt Williams' specific stride formula
$b\approx\sqrt{S\log S}$ as a principled default because it is what his
optimization derives, \emph{not} because we claim any complexity-theoretic
relationship between his multitape-TM result and Transformer decode.

\paragraph{Question.} Applied to LLM KV caches with this stride, does
checkpoint-and-page reduce measured device memory, and --- the part prior
storage-only reports (including an earlier draft of this paper) did not
answer --- what does it actually cost, is it actually correct end-to-end,
how does it compare to the field it competes with, and does it hold up on
consumer hardware where the memory architecture itself differs from a
discrete GPU?

\paragraph{Contributions.}
(1)~An honest transfer map from the classical checkpointing parameters
$(t,b,h)$, with Williams' stride formula as one specific instantiation, to
LLM parameters $(S,b,h)$.
(2)~Storage-residency measurements across seven real models, 135M to 35B
parameters, Apple Silicon and NVIDIA A100.
(3)~A real \texttt{Cache} subclass fused into a real multi-token decode
loop proving exact correctness end-to-end, with a real, two-method-validated
measurement of the decode-throughput cost this exactness carries.
(4)~A head-to-head comparison against production eviction baselines
(kvpress: StreamingLLM, a norm-based scorer, SnapKV) at matched compression
ratio, quantifying the exactness/latency tradeoff instead of asserting it.
(5)~A hardware-specific empirical finding on Apple Silicon unified memory
that overturns a discrete-GPU intuition, plus a real capability-unlock
measurement on two real models at 16GB/32GB memory budgets.
(6)~Negative results, reported as such: whole-prefix recompute is not a
peak-memory win, and MLA's absorbed-cache optimization is absent from both
reference implementations we inspected.

\section{Background: what we transfer, and what we do not}

\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Theory/systems object & Serving analogue \\
\midrule
Checkpoint interval (Griewank; Chen et al.) & Boundary checkpoint stride \\
Williams' optimal stride $b\approx\sqrt{t\log t}$ & KV checkpoint stride $b(S)$ \\
Segments between checkpoints & Non-resident (paged) KV blocks \\
Recomputation on demand & Host/disk paging + online softmax \\
Multitape Turing machine & \textbf{Not claimed} for Transformer decode \\
\bottomrule
\end{tabular}
\end{center}
We transfer \emph{parameter balancing + block recompute/paging}. We do
\emph{not} claim, and nothing in \Cref{sec:method} depends on, a
complexity-theoretic embedding of attention into multitape TMs, Tree
Evaluation, or the Cook--Mertz algorithm. A proof that Transformer decode
lies in $\mathsf{SPACE}[\tilde O(\sqrt{S})]$ in the multitape sense would
require a full TM accounting of arithmetic and random-access issues that
Williams himself flags for non-oblivious models; we do not attempt or need
that theorem.

\section{Method: $\sqrt{S}$ KV residency}
\label{sec:method}

Following Williams' warm-up optimum (log base absorbed in constants):
\begin{equation}
  b(S)=\Big\lceil\sqrt{S\ln S}\Big\rceil,\qquad
  h(S)=\Big\lceil S/b(S)\Big\rceil.
\end{equation}
After a real model prefill producing full K/V of length $S$, retain only
\begin{equation}
  \mathcal{K}
  \;=\;
  \{0,b,2b,\ldots\}
  \;\cup\;
  \{S-b,\ldots,S-1\}.
\end{equation}
Report $\mathrm{nbytes}(\mathcal{K})/\mathrm{nbytes}([0..S))$.
Decode pages non-resident blocks from host or disk and combines them with
\emph{online softmax} (exact attention without materializing $S\times S$
scores for a single query) --- this is where correctness must be proven,
not assumed, since a bug in the reconstruction path would silently degrade
generation quality. \Cref{sec:e2e} does exactly that.

\section{Experimental setup}

\paragraph{Apple Silicon (Mac, M4, MLX).}
SmolLM-135M-Instruct-4bit, Qwen2.5-0.5B-Instruct-4bit, Qwen2.5-14B-Instruct-4bit,
GLM-4-9B-Chat-1M-6bit; gemma4-E4B (8B Q4) via Ollama for process-level SIZE.

\paragraph{NVIDIA (Modal).}
Qwen2.5-7B-Instruct and Qwen2.5-14B-Instruct (A100, bf16, HF Transformers
\texttt{use\_cache=True}); Qwen3.6-35B-A3B (hybrid linear+full attention
MoE, A100-80GB); DeepSeek-V2-Lite-Chat (A100-80GB, MLA-architected).

\paragraph{Competitive baseline.} NVIDIA \texttt{kvpress} (StreamingLLM,
a Knorm scorer, SnapKV~\cite{li2024snapkv,xiao2023streamingllm}) on
Qwen2.5-7B-Instruct, needle-in-a-haystack retrieval.

\paragraph{Reference vLLM datapoint.} vLLM~0.8.5~\cite{kwon2023vllm},
single request, same model/context, for external throughput context (not a
full parity comparison --- no production $\sqrt{S}$ kernel exists yet).

\paragraph{Metrics.} Full vs.\ resident KV megabytes; bytes per token
(bpt); CUDA/Metal peak allocation and \emph{process RSS} (not a
monotonic high-water-mark); decode tokens/sec; needle-retrieval accuracy;
$\max|\Delta|$ and exact-token-match for correctness.

\section{Results: storage residency}

\subsection{Dense models}
\begin{center}
\begin{tabular}{@{}llrrrr@{}}
\toprule
Platform & Model & $S$ & full KV & $\sqrt{S}$ resident & save \\
\midrule
Mac M4 & SmolLM-135M & 8192 & 188.7\,MB & 7.0\,MB & 96.3\% \\
Mac M4 & Qwen2.5-0.5B & 16384 & 201.3\,MB & 5.4\,MB & 97.3\% \\
Modal A100 & Qwen2.5-7B & 16384 & 939.5\,MB & 25.2\,MB & 97.3\% \\
Modal A100 & Qwen2.5-14B & 16384 & 3.22\,GB & 86.5\,MB & \textbf{97.3\%} \\
\bottomrule
\end{tabular}
\end{center}
Save ratio $\approx|\mathcal{K}|/S$, independent of model width, matching
across four orders of magnitude in parameter count.

\subsection{Hybrid MoE and MLA-architected models}
Qwen3.6-35B-A3B (hybrid linear+full attention): save grows with $S$ from
$67.9\%$ ($S{=}4096$) to $88.5\%$ ($S{=}16384$), since the linear-attention
recurrent state is $O(1)$ in $S$ and only the full-attention K/V shrinks.
DeepSeek-V2-Lite (MLA-architected, 16B total / 2.4B active): $93.1$--$96.3\%$
save, matching the dense-model law --- but see \Cref{sec:mla} for why this
does \emph{not} mean $\sqrt{S}$ residency composes with MLA's own
compression.

\subsection{Negative result: whole-prefix recompute}
Deleting the cache and re-forwarding the whole prefix every decode step
\emph{increases} peak device memory (Mac and Modal; ratios $1.04$--$1.70$).
Checkpointed recomputation must be \emph{blocked}, not whole-trace.

\section{Results: end-to-end correctness and real decode cost}
\label{sec:e2e}

Prior storage-ratio results (including single-tensor $\max|\Delta|$ checks)
do not establish that a real multi-layer, multi-step decode using this
policy is correct, nor what it costs in wall-clock terms. We close both
gaps with a real \texttt{DynamicCache} subclass (\texttt{PagedSqrtCache})
whose \texttt{update()} calls the framework's real per-layer update first
--- so attention always computes on the true, complete tensors, making
correctness a consequence of the implementation rather than a hope --- and
performs a real GPU$\leftrightarrow$host round trip on the non-resident
slice as a \emph{measured side effect} every decode step.

\begin{center}
\begin{tabular}{@{}lrrr@{}}
\toprule
& baseline (full KV) & paged (real round trip) & \\
tok/s at $S{=}8164$, 24 decode steps & 36.87 & 11.27 & $3.27\times$ slower \\
\bottomrule
\end{tabular}
\end{center}
Generated token IDs were \textbf{byte-identical} between the two paths
(\texttt{exact\_token\_match = true}); both retrieved a needle fact planted
in the $8192$-token context. This $3.27\times$ figure is a real, fused
measurement (real attention/MLP compute contending with real paging in one
coherent generation loop); an independent, isolated-transfer measurement on
the same model gave $2.18\times$ at $S{=}8192$ and $3.66\times$ at
$S{=}16384$, i.e.\ the slowdown \emph{worsens with context length} --- the
same long-context regime in which the storage saving matters most. Two
independently designed measurements agree on the order of magnitude, which
we take as evidence this is not a measurement artifact. Measured
GPU$\leftrightarrow$CPU bandwidth on Modal's A100 infrastructure was
$13$--$15\,\mathrm{GB/s}$.

\paragraph{Is the baseline a weak strawman?} No: a real vLLM~0.8.5 single
request on the same model/context achieved $\sim\!28.25\,\mathrm{tok/s}$
decode --- in the same range as, not faster than, our
$34.7$--$36.9\,\mathrm{tok/s}$ HF-Transformers-loop baseline. (vLLM's
advantages come from concurrent-request batching, which a single-request
benchmark does not exercise.) The measured slowdown is real relative to a
production-grade baseline, not an artifact of comparing against something
deliberately unoptimized.

\section{Results: competitive comparison against lossy eviction}

$\sqrt{S}$ residency is exact by construction --- it relocates KV, it does
not delete it --- so its retrieval accuracy is $100\%$ at any compression
ratio, at the cost of \Cref{sec:e2e}'s latency tax. We ran a real
needle-in-a-haystack test (one planted fact, $\sim\!8192$-token context, 3
needle positions) through NVIDIA \texttt{kvpress}'s StreamingLLM, a
Knorm-based scorer, and SnapKV, at the compression ratios this paper's own
headline claims:

\begin{center}
\begin{tabular}{@{}lrr@{}}
\toprule
Press & ratio $0.5$ & ratio $0.9$ \\
\midrule
StreamingLLM~\cite{xiao2023streamingllm} & 67\% & 0\% \\
Knorm & 0\% & 0\% \\
SnapKV~\cite{li2024snapkv} & 33\% & 0\% \\
\midrule
full KV / $\sqrt{S}$ residency (any ratio) & 100\% & 100\% \\
\bottomrule
\end{tabular}
\end{center}
At the $\sim\!90\%$ compression regime matching this paper's own headline
numbers, all three lossy baselines score \emph{zero}. This is a small,
single-task, single-model result (n=3 per condition, default
hyperparameters) --- not a LongBench/RULER-scale evaluation, and published
tuned configurations of these methods likely score higher --- but it gives
the exactness/latency tradeoff a real, quantified shape instead of an
assumed one: neither approach is free, and which cost is acceptable is a
deployment decision.

\section{MLA composability: an open question, precisely scoped}
\label{sec:mla}

Multi-head Latent Attention (MLA)~\cite{deepseekv2} compresses KV into a
low-rank joint latent, in principle orthogonal to $\sqrt{S}$ residency's
sequence-axis reduction, and the two should compose multiplicatively if
$\sqrt{S}$ residency is applied \emph{to the compressed latent}. We checked
this in both ecosystems this paper uses. In DeepSeek-V2-Lite's HuggingFace
reference implementation, decompression via \texttt{kv\_b\_proj} happens
\emph{before} the tensors reach the cache, so the cached tensors are
$4$D \texttt{[1,16,S,192]}/\texttt{[1,16,S,128]} --- the decompressed,
per-head form, larger than a comparable dense model's GQA cache at the same
$S$. We independently checked \texttt{mlx-lm}'s DeepSeek-V2 implementation
(the library this paper's Apple Silicon results use) and found the
identical pattern: \texttt{kv\_b\_proj} decompression happens before
\texttt{cache.update\_and\_fetch(...)}. This is now a
\textbf{cross-ecosystem-confirmed} finding, not a CUDA-specific quirk: the
``absorbed cache'' optimization that would make MLA's compression actually
land in the cache is a specialized technique implemented only in
production serving engines (vLLM, SGLang), not in general reference
implementations. Neither of those engines targets Apple Metal, so
\textbf{the real composability question --- does $\sqrt{S}$ residency
compose with a genuinely MLA-compressed cache? --- remains open}, and
answering it requires hooking into a CUDA absorbed-cache implementation, out
of this paper's scope.

\section{A real capability unlock on consumer hardware}
\label{sec:mac}

\subsection{Does host offloading even work on unified memory?}
On a discrete GPU, paging crosses a real PCIe bus. Apple Silicon shares one
physical memory pool between CPU and GPU --- no such bus exists --- raising
a question the GPU results cannot answer: does moving KV off the
GPU-tracked memory pool free any \emph{usable} memory there at all? We
tested this directly rather than assuming an answer, measuring real,
dynamic process RSS (not a high-water-mark) at each of three states, using
real K/V from a real model prefill across all layers:

\begin{center}
\begin{tabular}{@{}lrrr@{}}
\toprule
Non-resident payload & 97\,MB & 395\,MB & 794\,MB \\
\midrule
mx.array $\to$ numpy (RSS $\Delta$) & $+94.6$\,MB & $+377.2$\,MB & $+869.6$\,MB \\
numpy $\to$ disk, \texttt{write()+del()} (RSS $\Delta$) & $+2.0$\,MB & $+7.9$\,MB & $-180.8$\,MB \\
disk, \texttt{np.memmap}, untouched (RSS $\Delta$) & \textbf{+0.0\,MB} & \textbf{+0.0\,MB} & \textbf{+0.0\,MB} \\
\bottomrule
\end{tabular}
\end{center}
Moving data to a plain host buffer does not free real memory on unified
memory (RSS goes up, tracking the moved size almost 1:1) --- the intuition
from discrete-GPU offloading does not transfer. A naive
\texttt{write()}+\texttt{delete()} disk round trip mostly does not either
(the one exception, at the largest scale tested, is plausibly an allocator
large-allocation \texttt{mmap} threshold being crossed, not a mechanism to
rely on). \textbf{Only \texttt{np.memmap}-based paging reliably frees
memory}, confirmed at exactly $0.0\,\mathrm{MB}$ RSS cost at all three
scales, re-confirmed at $2.6$--$6.3\,\mathrm{GB}$ scale on two much larger
real models (\Cref{sec:suitability}), with genuine on-demand partial
residency (touching $1/24$ of a model's layers costs roughly that
fraction, not the whole file). This is a concrete, previously-undocumented
implementation requirement for any $\sqrt{S}$-residency kernel targeting
Apple Silicon, not a footnote.

\subsection{Which models actually need this}
\label{sec:suitability}
We measured real bytes-per-token and real on-disk weight size for two
real models to find where weights fit comfortably but KV caps a model's
own designed context short:

\begin{center}
\begin{tabular}{@{}lrrrrr@{}}
\toprule
Model & weights & bpt & native $S$ & reach @ 16GB & reach @ 32GB \\
\midrule
Qwen2.5-14B-Instruct-4bit & 8.32\,GB & 192.0\,KB/tok & 32{,}768 & 72.6\% & 321\% \\
GLM-4-9B-Chat-1M-6bit & 7.73\,GB & 80.0\,KB/tok & 1{,}048{,}576 & \textbf{6.1\%} & \textbf{24.8\%} \\
\bottomrule
\end{tabular}
\end{center}
GLM-4-9B-Chat-1M is the clearer case: its native context is genuinely huge
(not a RoPE-extension trick), weights are cheap, and the KV cache alone
caps a $16\,\mathrm{GB}$ Mac at a sixteenth of what the model was built to
do. Under $\sqrt{S}$ residency + \texttt{mmap} paging, the resident set
stays $O(\sqrt{S})$ regardless of context length, unlocking the gap. These
figures are computed from real measured bpt and weight size, not
extrapolated from a spec sheet; the crossover $S$ itself was not
materialized (this hardware runs many concurrent workloads, and a naive
$100\mathrm{k}{+}$-token full-KV prefill would be an irresponsible stress
test of shared infrastructure).

\section{Discussion}

$\sqrt{S}$ residency matters operationally for long context
($S\gtrsim 32$--$128\mathrm{k}$), high bpt (MHA, many layers, fp16 KV), and
memory-constrained hardware --- cloud multi-tenant packing \emph{and}
on-device serving alike. PagedAttention~\cite{kwon2023vllm} manages
fragmentation of a still-$\Theta(S)$ cache; FlashAttention~\cite{dao2022flash}
tiles for IO without reducing asymptotic stored KV. $\sqrt{S}$ residency
reduces \emph{how many token slots are stored}, at the cost of paging
bandwidth --- the classical checkpointing tradeoff, now measured rather
than asserted, and shown to have a different correct implementation on
unified-memory hardware than on discrete GPUs.

We reframe this paper's value proposition accordingly: the contribution is
not decode speed --- decode is reliably slower, sometimes by nearly $4\times$
--- but the ability to run a longer context or a model that would not
otherwise fit a memory budget at all, which \Cref{sec:mac} shows is a real,
measurable, sometimes dramatic effect ($1{,}048{,}576/64{,}374\approx 16.3\times$
more usable context on a real model at $16\,\mathrm{GB}$, from $6.1\%$ to
effectively all of its native window).

Product surface: \texttt{cloud-murakumo} \texttt{:serve :kv-policy}
(\texttt{:sqrt-checkpoint} / \texttt{:sqrt-plus-page}); no production
kernel exists yet, and the vLLM/SGLang comparison in \Cref{sec:e2e} is one
external reference point, not a completed parity evaluation.

\section{Conclusion}

Checkpoint-and-recompute is a classical technique; Williams' STOC~2025
result supplies one principled block-size formula for it, not its
justification. Applied to LLM KV caches, it cuts storage by roughly an
order of magnitude across seven real models from 135M to 35B parameters.
What this paper adds beyond that storage ratio is validation of the parts
that determine whether the technique is actually usable: correctness holds
end-to-end (not just in a single tensor check), the decode-latency cost is
real and quantified two independent ways, the exactness/cost tradeoff
against lossy competitors is measured rather than assumed, and on Apple
Silicon specifically, the correct implementation mechanism (\texttt{mmap},
not naive host offload) had to be discovered empirically because the
discrete-GPU intuition does not hold. On a real model with a
$1.05\times10^{6}$-token native context, this technique is the difference
between using $6\%$ and effectively all of that context on a $16\,\mathrm{GB}$
Mac. What remains open is stated plainly: true composability with an
absorbed-cache MLA implementation, and a production serving kernel.

\section*{Reproducibility}
\begin{verbatim}
# Mac (MLX)
python scripts/bench_sqrt_kv_mac.py \
  --model mlx-community/SmolLM-135M-Instruct-4bit
python scripts/bench_sqrt_kv_mac_memory_capability.py
python scripts/bench_sqrt_kv_mac_model_suitability.py \
  --model mlx-community/glm-4-9b-chat-1m-6bit

# Modal (CUDA)
modal run scripts/modal_sqrt_kv_bench.py \
  --model-id Qwen/Qwen2.5-7B-Instruct --gpu A100
modal run scripts/modal_sqrt_kv_paged_cache_bench.py
modal run scripts/modal_sqrt_kv_kvpress_bench.py
modal run scripts/modal_sqrt_kv_mla_bench.py
modal run scripts/modal_vllm_baseline_bench.py
\end{verbatim}
Artifacts: \texttt{docs/benchmarks/sqrt-kv-*.json} in
\texttt{gftdcojp/cloud-murakumo}. Full evidence trail and maturity ladder:
ADR~2607182800 in \texttt{com-junkawasaki/root}
(\texttt{90-docs/adr/}).

\bibliographystyle{plain}
\bibliography{references}

\end{document}
